Learning complex models with invertible neural networks: a likelihood-free Bayesian approach

### 2019-07-18 latex/paper_main.tex
Add seqRNA example

\title{Learning complex stochastic models with invertible neural networks: a likelihood-free approach to Bayesian inference}

@@

The \textit{summary network} is responsible for learning the most informative summary statistics directly from data. It should be designed to follow the functional and probabilistic symmetries inherent in the data, e.g. a permutation invariant network for \textit{i.i.d.} data \cite{bloem2019probabilistic}, a recurrent network \cite{gers1999learning} or a convolutional network \cite{long2015fully} for data with temporal or spatial dependencies. Ideally, it should also be independent of the number of observations. The computation of summary statistics is a crucial aspect in likelihood-free inference. Previous approaches typically used hand-crafted summary statistics tailored to the specific application \cite{raynal2018abc, sunnaaker2013approximate}. However, in many applications, finding an optimal set of summary statistics is far from obvious. The summary network eliminates the need to manually decide on and compute a fixed set of summary statistics and, thus, makes the method independent of the format or the size of the data.

@@

As a last example for our method, we use simulations from the \textit{Splat} model to recover the data-generating parameters. This is a challenging task for any Bayesian method, as the data-generating mechanism is a complicated algorithm with no available likelihood for joint estimation of the parameters. Moreover, the observed data exhibits multiple local and non-local dependencies. We simulate $200\ 000$  \textit{Gene x Cell} count matrices with different parameter settings. Each entry in a count matrix contains the number of times a gene has been sequenced in a different cell. We simulate counts for $40$ genes and $100$ cells, that is, each $\boldsymbol{X} \in \mathbb{R}^{40\times100}$. The parameters considered for estimation, along with a short description, are listed in \autoref{table:Table 1}.

The parameter estimation task is thus to recover $\boldsymbol{\theta} = (\alpha,\beta,\mu^{L},\sigma^{L},\pi^{O},\mu^{O},\sigma^{O},\phi)$ from an observed count matrix $\boldsymbol{X}$. We train the networks for 30 epochs through the entire dataset and evaluate the performance on a separate validation set of $300$ count matrices. To stabilize training and avoid exploding gradients due to large counts in the input matrices, we apply a log transformation to each $\boldsymbol{X}$. Simulating the data took approx. $8$ hours. Training the networks took approx. one day with the reference table approach. Inference on $300$ datasets with $2000$ posterior samples per parameter took approx. $1.7$ seconds. 
\begin{figure}[H]
\centering
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_plot.png}
    \caption{The \textit{Splat} simulation diagram. Adapted after \cite{zappia2017splatter}.}
    \label{fig:Fig.7a}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_scatter.png}
    \caption{Parameter recovery}
    \label{fig:Fig.7b}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_sbc.png}
    \caption{Simulation-based calibration (SBC)}
    \label{fig:Fig.7c}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_density.png}
    \caption{Example posterior samples}
    \label{fig:Fig.7d}
\end{subfigure}

\caption[short]{Results on the \textit{Splat} scRNA simulation model. \textbf{(a)} The \textit{Splat} simulation diagram; \textbf{(b)} Parameter recovery; \textbf{(c)} Simulation-based calibraton. The SCB plots indicate a slight underestimation of the true posterior variance.  \textbf{(d)} Example full parameter posteriors from three datasets.} \label{fig:Fig.7}
\end{figure}
The results on the scRNA example are depicted in \autoref{fig:Fig.7}. First, recovery of most parameters is very good, with maximum $R^{2}$ of $0.98$ and minimum NRMSE of $0.039$ for the library location parameter $\mu^{L}$. Interestingly, the rate parameter $\beta$ appears to be nearly unrecoverable, as indexed by the scatterplots in \autoref{fig:Fig.7b}. Furthermore, the estimated posterior of $\beta$ is often bimodal, covering the whole prior range (\autoref{fig:Fig.7c}). This might indicate that the parameter is not identifiable, or that some multivariate trade-offs exist in the parameter space of the \textit{Splat} model. The latter is also confirmed by observing a few marked outliers in the scatterplots of $\alpha$, $\sigma^{O}$, and $\pi^{O}$. Finally, the SCB histograms indicate that our method tends to slightly underestimate the true posterior variance of the $\alpha$, $\beta$, $\mu^{O}$ $\pi^{O}$, and $\sigma^{O}$ parameters. This issue might be due to the fact that the networks were trained with the reference-table approach and might disappear if the online learning approach is employed.

>>>>

\title{Learning complex stochastic models with invertible neural networks: a new approach to Bayesian inference}

@@

The \textit{summary network} is responsible for learning the most informative summary statistics directly from data. It should be designed to follow the functional and probabilistic symmetries inherent in the data, e.g. a permutation invariant network for \textit{i.i.d.} data \cite{bloem2019probabilistic}, a recurrent network \cite{gers1999learning} or a convolutional network \cite{long2015fully} for data with temporal or spatial dependencies. Ideally, it should also be independent of the number of observations. The computation of summary statistics is a crucial aspect in likelihood-free inference. Previous approaches typically used hand-crafted summary statistics tailored to the specific application \cite{raynal2018abc, sunnaaker2013approximate}. However, in many applications, finding an optimal set of summary statistics is far from obvious. This also applies to the models considered in this work. The summary network eliminates the need to manually decide on and compute a fixed set of summary statistics and, thus, makes the method independent of the format or the size of the data.

@@

As a last example for our method, we use simulations from the \textit{Splat} model to recover the data-generating parameters. This is a challenging task for any Bayesian method, as the data-generating mechanism is a complicated algorithm with no available likelihood for joint estimation of the parameters. Moreover, the observed data exhibits multiple local and non-local dependencies. We simulate $250\ 000$  \textit{Gene x Cell} count matrices with different parameter settings. Each entry in a count matrix contains the number of times a gene has been sequenced in a different cell. We simulate counts for $40$ genes and $100$ cells, that is, each $\boldsymbol{X} \in \mathbb{R}^{40\times100}$. The parameters considered for estimation, along with a short description, are listed in \autoref{table:Table 1}.

The parameter estimation task is thus to recover $\boldsymbol{\theta} = (\alpha,\beta,\mu^{L},\sigma^{L},\pi^{O},\mu^{O},\sigma^{O},\phi)$ from an observed count matrix $\boldsymbol{X}$. We train the networks for 30 epochs through the entire dataset and evaluate the performance on a separate validation set of $300$ count matrices. To stabilize training and avoid exploding gradients due to large counts in the input matrices, we apply a log transformation to each $\boldsymbol{X}$. Simulating the data took approx. $8$ hours. Training the networks took approx. one day with the reference table approach. Inference on $300$ datasets with $2000$ posterior samples per parameter took approx. $1.7$ seconds. To our knowledge, this is the first Bayesian treatment of the \textit{Splat} model.
\begin{figure}[H]
\centering
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_plot.png}
    \caption{The \textit{Splat} simulation diagram. Adapted after \cite{zappia2017splatter}.}
    \label{fig:Fig.7a}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_scatter.png}
    \caption{Parameter recovery}
    \label{fig:Fig.7b}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_sbc.png}
    \caption{Simulation-based calibration (SBC)}
    \label{fig:Fig.7c}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_density.png}
    \caption{Example posterior samples}
    \label{fig:Fig.7d}
\end{subfigure}

\caption[short]{Results on the \textit{Splat} scRNA simulation model. \textbf{(a)} The \textit{Splat} simulation diagram; \textbf{(b)} Parameter recovery; \textbf{(c)} Histograms of the rank statistic for simulation-based calibration (SCB); \textbf{(d)} Example full parameter posteriors from three datasets.} \label{fig:Fig.7}
\end{figure}
The results on the scRNA example are depicted in \autoref{fig:Fig.7}. First, recovery of most parameters is very good, with maximum $R^{2}$ of $0.999$ and minimum NRMSE of $0.01$ for the library location parameter $\mu^{L}$. Interestingly, the rate parameter $\beta$ appears to be unrecoverable, as indexed by the scatterplots in \autoref{fig:Fig.7b}. Furthermore, the estimated posterior of $\beta$ is approximately uniform, covering the entire prior range (\autoref{fig:Fig.7c}). This indicates that no information about the parameter is present in the data. We also observe some multivariate trade-offs in the parameter space suggested by the marked outliers in the scatterplots of $\mu^{O}$ and $\sigma^{O}$. Finally, the SCB histograms indicate no systematic bias in the approximate posteriors across all parameters. The posterior mean of $\alpha$ appears to be slightly overestimated, whereas the opposite trend is present for $\pi^{O}$ and $\sigma^{L}$. These inaccuracies might be due to the fact that the networks were trained with the reference-table approach and might disappear by employing the online learning approach.

### 2019-07-18 latex/paper_supplementary.tex
Add seqRNA example

\subsubsection*{Summary network}
We use the recently introduced permutationally invariant neural network \cite{bloem2019probabilistic} for the \textit{i.i.d.} regression data. This network architecture is a good choice for this example, as each observation is independent of all other, and thus all permutations of a given dataset should lead to the same parameter estimates. Furthermore, invariant networks are independent of the number of observations.

@@

@@

\subsubsection*{Summary network}
We use a permutationally invariant neural network \cite{bloem2019probabilistic} for the \textit{i.i.d.} reaction times (RT) data. Similarly to the toy Regression example, each response in an RT dataset is assumed to be independent of all others, so permutations of the dataset must lead to the same parameter estimates.

@@

@@

@@

@@

\movie{Type caption for the movie here.}

>>>>

\subsubsection*{Summary network}
We use the recently introduced permutation invariant neural network \cite{bloem2019probabilistic} for the \textit{i.i.d.} regression data. This network architecture is a good choice for this example, as each observation is independent of all other, and thus all permutations of a given dataset should lead to the same parameter estimates. Furthermore, permutation invariant networks are independent of the number of observations.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Ricker_plot_multiple.png}
\caption{Some example Ricker time-series generated with different parameters drawn from the prior.} 
\end{figure}

@@

\subsubsection*{Summary network}
We use a permutation invariant neural network \cite{bloem2019probabilistic} for the \textit{i.i.d.} reaction times (RT) data. Similarly to the toy Regression example, each response in an RT dataset is assumed to be independent of all others, so permutations of the dataset must lead to the same parameter estimates.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Levy_plot_multiple.png}
\caption{Some example Lévy-Flight datasets generated with different parameters drawn from the prior.} 
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{SIR_plot_multiple.png}
\caption{Some example SIR time-series generated with different parameters drawn from the prior.} 
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Splatter_plot_multiple.png}
\caption{A heatmap of a single $Genes \star Cells$ count matrix generated with different parameters drawn from the prior.} 
\end{figure}

@@
